---
title: "Movie Network"
author: "Kenny Darrell"
date: "September 25, 2014"
output: html_document
---

I just created an account on [shinyapps](https://www.shinyapps.io/). It seems pretty cool to be able to launch an app that can be hosted more globally. Taking a break from my last few posts about temporal data and jump back to one of my other interests, graph data. I used the graph of [Pulp Fiction](http://moviegalaxies.com/movies/660-Pulp-Fiction) from [moviegalaxies.com](moviegalaxies.com). I thought it would be cool to create an interactive application using [d3](http://d3js.org/) and [Shiny](http://shiny.rstudio.com/).

```{r, eval = FALSE}
require(rgexf)
require(httr)
library(XML)
options(stringsAsFactors = FALSE)

# Site of movie files.
home <- 'moviegalaxies.com/'
# Directory on server of gexf files.
g <- 'http://media.moviegalaxies.com/gexf'

# Init structure to store movie info.
info <- list()

# Loop over all movies, hack becuase I have no idea how to start,
# the have names that are random numbers from 1 to 914.
for (i in 1:925) {
  
  html <- strsplit(as.character(GET(paste(home, i, sep = ''))), '\n')[[1]]
  
  title <- grep('og:title', html, value = T)[1]
  title <- gsub('\" /> ', '', strsplit(title, ' - ')[[1]][2])
  gex <- grep('com/gexf', html, value = T)
  gex <- strsplit(gex, 'gexf')[[1]][2]
  gex <- paste(g, gex, 'gexf', sep = '')
  misc <- grep('keywords', html, value = T)[1]
  
  # There is an issue with 'Face/Off' so take out /
  if (i == 297) title <- 'Face Off'
  if (title != '') {
    # Get the gexf file from the site an place it into a file
    gex <- as.character(GET(gex))
    cat(gex, file = paste('gex/', title, '.gexf', sep = ''))
    
    info[[title]] <- misc
  }
  cat(i, title, '\n')
}



# Once you have the value and the movie name scrapped from each site that
# is actually a movie, pull that file from the server.
ig <- list()

for (i in names(info)) {
  print(i)
  # This file is corrupt
  if (i != "12") {
    ig[[i]] <- gexf.to.igraph(read.gexf(paste('gex/', i, '.gexf', sep = '')))
  }
}

info <- info[-2]

save(ig, info, file = 'movies.RData')

```

You can also embed plots, for example:

```{r, eval=FALSE}
library(igraph)
library(d3Network)
#library(rgexf)
#.libPaths('/home/likewise-open/CHO/kdarrell/R/x86_64-pc-linux-gnu-library/3.1')

igraph_2_d3 <- function(igr) {
  Source <- c()
  Target <- c()
  
  x <- V(igr)
  for (i in seq(length(E(igr)))) {
    e <- get.edge(igr, i)
    
    s <- if(is.null(x[e[1]]$name)) x[e[1]] else x[e[1]]$name
    t <- if(is.null(x[e[2]]$name)) x[e[2]] else x[e[2]]$name
    Source <- c(Source, s)
    Target <- c(Target, t)
  }
  
  data.frame(Source, Target)
}

load('info.RData')
load('ig.RData')


clean <- function(x) {
  x <- gsub('<link href=\"/images/favicon.ico\" type=\"image/x-icon\" rel=\"icon\"', '', x)
  x <- gsub(' /><link href=\"/images/favicon.ico\" type=\"image/x-icon\" ', '', x)
  x <- gsub('rel=\"shortcut icon\" /><meta name=\"keywords\"', '', x)
  x <- gsub(' content=\"social graph, movies, social network, moviegalaxies, ', '', x)
  x <- gsub('\" />', '', x)
  x <- gsub(',', ',<br>', x)
  x
}

shinyServer(function(input, output) {
  output$networkPlot <- renderPrint({
    #x <- gexf.to.igraph(read.gexf(paste('gex/', input$x, '.gexf', sep = '')))
    #cat(info[[input$x]])
    cat(clean(info[[which(names(info) == input$x)]]))
    d3SimpleNetwork(igraph_2_d3(ig[[input$x]]), 
                    width = 500, height = 450,
                    textColour = 'orange', linkColour = 'red',
                    parentElement = '#networkPlot')
  })
})
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r, eval = F}
library(shiny)
load('info.RData')

shinyUI(fluidPage(
  tags$head(
    tags$script(src = 'd3.min.js')
  ),  
  
  titlePanel("Movie Network"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput('x','Select Movie', setdiff(names(info), "12"))
    ),
  
  mainPanel(
   # includeHTML("d3.js"),
    textOutput('txt', container = div),
    htmlOutput('networkPlot')
  )
 )
))
```
