---
title: '2048'
author: "Kenny Darrell"
date: "July 1, 2014"
output: html_document
---

Have you ever played 2048? Some months back in the winter I bacame extremely addicted to this game. All sorts of posts on Hacker News had new implementations of the game. Some that mearly changed the from numbers to [images](http://doge2048.com/) all the way to those that had there own unique [method](http://cesarkawakami.github.io/2048-5D/). You can find a ton of them [here](http://get2048.com/).

I was able to beat it many times and got close to 4098 a few times. I really became interested in what was the best stratagy to use. Many people created AI to play perfectly. I started thinking how good is randomly hitting directions. I could not find one that had and API to feed random directions into so I coded my own. I think now there may be but I did this a few months ago and just wanted to write about it.

You can get my implentation by running the following in R.

```{r, cache=TRUE}
devtools::source_url('http://darrkj.github.io/home/rcode/r2048.R')
```

I am just going to run this a lot of times and see what the outcome is. I am using a very simple stratagy that chooses two directions more often than the other two. This seemed to be very effective. One method is uniform over the directions. There is also a very you cna play in the console but it is very ugly.

```{r, cache=TRUE}
runs <- 100

a <- c()
for (i in seq(runs)) {
  if (i %% 50 == 0) print(i)
  a <- rbind(a, auto_play(gen_board(), c(.25, .25, .25, .25)) )
}
a <- log2(a)

#table(a[, 2])
mean(a[, 1])
var(a[, 1])
mean(a[, 3])

b <- c()
for (i in seq(runs)) {
  if (i %% 50 == 0) print(i)      
  b <- rbind(b, auto_play(gen_board(), c(.45, .001, .099, .45)) )
}
b <- log2(b)

#table(b[, 2])
mean(b[, 1])
var(b[, 1])
mean(b[, 3])

# Why does one take longer than the other.
```
