---
title: "neoterror"
author: "Kenny Darrell"
date: "July 3, 2014"
output: html_document
---

Lat time I walked through the Social Balance problem on networks of terroroist orgaizations. One thing that was interesting is that after all o fthe data was collected the code code became very awkward. Trying to create algorithms and traversals on graphs in R is not natural. There is a really nice solution to this problem as well. a few posts back I wrote about creating some wrappers around Neo4j to use from R. There is now a nice package wraps much more functionality up in a very usable manner. This gets rid of the code that we to traverse the graph. We only have to worry about gettin gthe data into a Neo4j database.

Lets start the data that was a available last time. This means that we have just used R for scraping and cleaning task. This code was all pretty standard.


ADD  FILE  HERE

```{r}
options(stringsAsFactors = FALSE)
library(igraph)
load('terrorNet.rda')
```


First we need make sure we have an instance of Neo4j up and running. Without going into to many details you need to [download Neo4j](http://www.neo4j.org/download). Once you unzip you can use the commands below on a Mac.

```
cd neo4j-community-2.0.3
bin/neo4j start
```

If you use a broswer and navigate to `http://localhost:7474/` you can see if it is indeed up and running.


Next we need to load the RNeo4j package. To get this setup there are instruction [here](http://nicolewhite.github.io/RNeo4j/). Once it is loaded we have to create a connection to the database instance. 

```{r}

library(RNeo4j)

graph = startGraph("http://localhost:7474/db/data/")

```



There are a few steps of preperation needed, but these are minor. We need to create

```{r eval = F}
# Create the unique set of all orgainzations
nodes <- unique(rbind(data.frame(id = net$from, name = net$fromn),
                      data.frame(id = net$to,   name = net$ton)))

# These will hold pointers to the node in the database when we create edges.
neoNodes <- list()

# Loop through all nodes and insert into db, retaining pointer.
for(i in 1:nrow(nodes)) {
  neoNodes[[i]] = createNode(graph, label = "Terror", name = nodes$id[i], 
                           aka = nodes$name[i]) 
}

# Now we add each edge to the db as a relationship.
for (i in 1:nrow(net)) {
  # Get index for which nodes are in this edge.
  from <- which(nodes == net$from[i])
  to <- which(nodes == net$to[i])
  # Add to db.
  createRel(from]], net$conn[i], neoNodes[[to]], how = net$type[i])
}
```





```{r}
q <- 'MATCH (a)-[r:hates]->(b)<-[s:likes]-(c), (a)-[t:hates]->(c) 
      return a.aka, b.aka, c.aka;'
tri <- cypher(graph, query = q)
```


```{r eval = FALSE}
devtools::source_url('http://darrkj.github.io/rcode/d3plot.R')

triples <- rbind(data.frame(Source = tri[, 1], Target = tri[, 2]),
           data.frame(Source = tri[, 1], Target = tri[, 3]),
           data.frame(Source = tri[, 2], Target = tri[, 3]))


d3plot(triples, 700, 600)
```




